\tit

### 2021-03-01 paper.tex
switching stars
vertical spacing in languages via predicate operations

@@

@@

@@

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.
For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
\agda{foldl}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

\figrefdef{predicates}{Predicate operations}{
\small
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{1.5ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from {\APT{Set} \AB ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{1ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).

>>>>

%% I think acmart minimizes these two.

@@

\setlength{\abovedisplayskip}{2ex}
\setlength{\belowdisplayskip}{2ex}
\setlength{\abovedisplayshortskip}{0ex}
\setlength{\belowdisplayshortskip}{0ex}

@@

In the definition of {\AB P \AF ☆}, \AB{concat} flattens a list of lists into a single list, defined as a right fold:

\agda{foldr-concat}

@@

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
\agda{foldl}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

\figrefdef{predicates}{Predicate operations}{
\small
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{1.5ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from {\APT{Set} \AB ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{1ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
\setstretch{2.5}
\agda{domain-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).
